### 2016-10-24 doc/publications/plos16/plos16.tex
some changes to plos16.tex

\section{Introduction}

The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion base pairs of the
human genome.

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
(SNPs) by applying linear or logistic regression analysis to each SNPs separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  Complex Trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are
not immune to overfitting as shown by ~\cite{Segal.2004} who gives examples of overfitting in the case where:
\begin{itemize}
\item a single tree is the most appropriate model; and
\item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
\end{itemize}
However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on each processor and combine the trees.

@@

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
Random forests this means that the the comutations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading is considered more performant. \marginpar{is performant a word?}
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

=======
%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

@@

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in CursedForest are an inherent property of Random
Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of CursedForest in
respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of a
model.

@@

Since the original implementation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant against overfitting. However it is not true that they will not overfit \cite[]{Segal.2004};
\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
    effective stragey for removing the problem.
  \end{enumerate}
\end{itemize}

@@

The current implementation of Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes}
%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data} 
Each dataset consists of n samples and m variables where n << m, and values for each variable are ordinal factors with three levels represented as numbers 0, 1, 2 (which correspond to the ?typical? encoding of genomics variation) randomly generated from uniform distribution with equal probabilities.  
The dichotomous response is generated from a two level Bernoulli trail with p (probability of success) calculated by applying inverse logit transformation to the continuous linear response (y) of five variables:

$$y = a_10* v_10 +  a_100*v_100 + a_1000*v_1000  +  a_10000*v_10000 + a_10000*v_100000  { where a_10 = 0.8, a_100 = 0.9, a_1000 = 1.1, a_10000 = 1.0, a_10000=1.15}$$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest selects previously demonstrated cancer variants }
%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
%variables for further analysis. 
As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by their importance to identify the variables predictive of the label or selecting the n most important variables for further analysis. 
We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset. 
We build this model on VCF files retrieved from the TCGA data portal. 
For the label, we use the cell type (i.e. tumor or normal).

>>>>

\section{Introduction}
\marginpar{change parallelisation to parallelization}

The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
human genome.

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
\marginpar{is Complex Trait a thing? why capitals?}
Complex Trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees.

@@

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genome data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading generally performs well. 
\marginpar{``performant'' -- a person who performs, an actor or celebrant }
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

@@

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in CursedForest are an inherent property of Random
Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of CursedForest in
respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

Since the original implementation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant against overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in the case where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.

\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
    effective stragey for removing the problem.
  \end{enumerate}
\end{itemize}

@@

The current implementation of Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes}
%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data}

Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a two level
Bernoulli trail with $p$ (probability of success) calculated by applying inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest selects previously demonstrated cancer variants }
%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
%variables for further analysis. 
As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by
their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).
